---
title: prompting
author: Johannes B. Gruber and Aleksandra Urman
date: today
format:
  html:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

# Introduction

This file does all the prompting for the article "Gender Representation in Large Language Models: A Cross-Linguistic and Cross-Model Analysis"

## Packages

We use Ollama with the R package rollama for the prompting.
To make the prompting more efficient and make sure errors can be easily recovered from, I use a custom version of `httr2::req_perform_parallel()`, which rollama uses under the hood.

```{r setup}
#| include: false
library(tidyverse)
library(rollama)

# the script can use several ollama servers and distribute requests between them
# options(rollama_server = c("http://localhost:11434", "http://192.168.2.45:11434"))

# special req perform with cache and load balancing
req_perform_parallel2 <- function(
  reqs,
  paths = NULL,
  on_error = "continue",
  progress = TRUE,
  skip_unfinished = FALSE,
  max_active = 3
) {
  exists <- file.exists(paths)
  resps <- vector("list", length = length(reqs))

  if (skip_unfinished) {
    resps[which(!exists)] <- NA_character_
    check <- map_lgl(paths[exists], check_json_response)
    paths[exists][!check] <- NA_character_
    return(resps)
  }

  if (!all(ping_ollama())) {
    stop("not all server are running")
  }

  check_model_installed(models)

  if (progress) {
    progress <- list(
      clear = TRUE,
      format = "Prompting {cli::pb_bar} | {cli::pb_percent} [{cli::pb_current}/{cli::pb_total} | rate: {cli::pb_rate} | ETA: {cli::pb_eta}]",
      format_done = "Done running {length(reqs)} prompts!"
    )
  }

  invisible(httr2::req_perform_parallel(
    reqs[!exists],
    paths = paths[!exists],
    on_error = on_error,
    progress = progress,
    max_active = max_active
  ))

  # verify that all responses can be parsed
  if (!all(map_lgl(paths, check_json_response))) {
    cli::cli_alert_warning("Some responses were corrupted, rerunning failed requests")
    # call itself to execute again
    resps <- req_perform_parallel2(
      reqs = reqs,
      paths = paths,
      on_error = on_error,
      # progress = progress,
      max_active = max_active
    )
  }
  return(paths)
}

# parse cached responses
parse_json_response <- function(r) {
  jsonlite::read_json(r) |>
    pluck("message", "content")
}

# make sure responses can be parsed
check_json_response <- function(json_file) {
  res <- try(jsonlite::read_json(json_file), silent = TRUE)
  if (methods::is(res, "try-error")) {
    cli::cli_warn("Corrupt json file will be deleted:  {json_file}")
    file.remove(json_file)
    return(FALSE)
  }
  if (pluck_exists(res, "error")) {
    cli::cli_warn(
      "Result contains error and will be deleted:  {json_file}.\n Error:\n{res$error}"
    )
    file.remove(json_file)
    return(FALSE)
  }
  if (!pluck_exists(res, "message", "content")) {
    cli::cli_warn("Content missing. file will be deleted: {json_file}")
    file.remove(json_file)
    return(FALSE)
  }
  return(TRUE)
}
start <- Sys.time() # note start time for later
```

## Data

We construct a grid of the different prompts (one for each language), the selected nouns, and a range of different models.

```{r models}
prompts <- read_csv("input/prompts.csv", col_types = "cc")
nouns <- read_csv("input/nouns.csv", col_types = "icccccccc") |>
  pivot_longer(
    cols = starts_with("Noun_"),
    names_to = "lang",
    values_to = "noun"
  ) |>
  mutate(lang = str_remove(lang, "Noun_")) |>
  rename(noun_id = id)
models <- c(
  "deepseek-r1:1.5b",
  "deepseek-r1:32b",
  "deepseek-r1:70b",
  "deepseek-r1:7b",
  "deepseek-r1:8b",
  "gemma:2b",
  "gemma:7b",
  "gemma3:12b",
  "gemma3:1b",
  "gemma3:4b",
  "gpt-oss:120b",
  "gpt-oss:20b",
  "llama3:8b",
  "llama3.1:70b",
  "llama3.1:8b",
  "llama3.2:1b",
  "llama3.2:3b",
  "llama4:latest",
  "mistral:latest",
  "mistral-small3.1:latest",
  "qwen2.5:0.5b",
  "qwen2.5:1.5b",
  "qwen2.5:3b",
  "qwen2.5:7b"
)
```


# Main

We construct the full set of prompts here.
To get only valid answers, we use the structured output technique, to limit the choices the model can make.
We also run each combination of prompt and model 10 times, to add variation.
In previous iterations, we noticed that Ollama does some caching of replies.
So me make sure to first run all prompts in iteration 1, then all prompts in iteration 2 and so on.

```{r prompts}
all_prompts <- nouns |>
  left_join(prompts, by = c("lang" = "version")) |>
  mutate(
    prompt = paste(prompt, noun, sep = "\n"),
    # schema for structured outputs. Limits the possible answers the models can give
    schema = map(answers, function(a) {
      a <- str_split(a, ";")[[1]]
      list(
        type = "object",
        properties = list(
          association = list(
            type = "string",
            enum = a
          )
        ),
        required = "association"
      )
    })
  ) |>
  expand_grid(model = models, seed = 1:5, temperature = seq(0, 1, 0.2)) |>
  mutate(
    prompt_id = tolower(paste(lang, noun, model, temperature, seed, sep = "_")),
    .before = 1L
  ) |>
  mutate(prompt_id = str_remove(prompt_id, "/")) |>
  # for temperature 0, the seed does not change the outcome, so we don't need it
  filter(!(temperature == 0 & seed > 1L)) |>
  # arrange by model to make sure no unnecessary loading and unloading of models happen
  # but also arrange by seed to not run same prompt + model back to back
  arrange(seed, temperature, model)
```

I use rollama to prepare the requests to the servers, which are then made with the custom req_perform_parallel2 function defined above.

```{r grid}
dir.create("cache", showWarnings = FALSE)
grid <- all_prompts |>
  mutate(
    req = pmap(
      list(prompt, model, schema, seed, temperature),
      function(p, m, s, sd, temp) {
        query(
          q = p,
          model = m,
          # we deactivated the structured output feature because thinking models (like gpt-oss) do not provide any answer at all
          # format = s,
          output = "httr2_request",
          model_params = list(seed = sd, temperature = temp)
        )[[1]]
      }
    )
  ) |>
  mutate(path = paste0("cache/", prompt_id, ".json"))
glimpse(grid)
```

In total, the grid contains `r nrow(grid)` requests to Ollama

```{r inference}
results <- grid |>
  mutate(
    resp = req_perform_parallel2(req, paths = path, skip_unfinished = FALSE)
  ) |>
  # remove when all prompting is done
  filter(!is.na(resp))
```

Results are written into a cache directory.
When a prompt has already been answered, we load the answer from disk instead of querying again.

```{r parsing}
results_parsed <- results |>
  mutate(answer = map_chr(resp, parse_json_response))
```

Quick glance at results:

```{r}
results_parsed |>
  count(lang, answer, sort = TRUE)
```

Note how much time was spent in total:

```{r time_spent}
time_spent <- results |>
  pull(resp) |>
  map(function(r) {
    res <- jsonlite::read_json(r)
    tibble(
      model = pluck(res, "model"),
      total_duration = pluck(res, "total_duration"),
      eval_duration = pluck(res, "eval_duration")
    )
  }) |>
  bind_rows()

time_spent_summary <- time_spent |>
  group_by(model) |>
  summarise(
    total_duration = sum(total_duration, na.rm = TRUE),
    eval_duration = sum(eval_duration, na.rm = TRUE)
  ) |>
  arrange(desc(eval_duration)) |>
  mutate(
    total_duration = make_difftime(total_duration / 1e9, units = "h"),
    eval_duration = make_difftime(eval_duration / 1e9, units = "h")
  )

time_spent_summary |>
  knitr::kable()
```

Since I always sent 10 requests at the same time, the `total_duration` is about 10x more than the eval duration -- as each query would sit in the queue until the last 9 queries have been processed by Ollama.
In total, all prompting took up `r sum(time_spent_summary$eval_duration)` of compute time.

# wrap up

save results:

```{r}
dir.create("data", showWarnings = FALSE)
saveRDS(results_parsed, "data/results_parsed.rds")
saveRDS(time_spent, "data/eval_time_spent.rds")
```

Some information about the session:

```{r}
sessionInfo()
Sys.time()
# note how long the script takes to (re-)run
Sys.time() - start
```
