---
title: 1._prompting.qmd
author: Johannes B. Gruber
date: today
format:
  html:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

# Introduction

A few words on what this document accomplishes.
I use one sentence per line to make it easier to see changes with version control, such as on GitHub.

- prompts
- word list

## Packages

Write which packages you use at the beginning so if a package is missing, the script fails early.

```{r}
#| include: false
library(tidyverse)
library(rollama)
# servers <- c("http://localhost:11434", "http://192.168.2.45:11434")
servers <- "http://192.168.2.29:11434"
if (!all(ping_ollama(servers, silent = TRUE))) {
  stop("not all server are running")
}
options(rollama_server = servers)
# special req perform with cache and load balancing
req_perform_parallel2 <- function(reqs,
                                  paths = NULL,
                                  on_error = "continue",
                                  progress = TRUE,
                                  max_active = 10) {
  exists <- file.exists(paths)
  resps <- vector("list", length = length(reqs))
  # pb <- list(
  #   clear = TRUE,
  #   format = "Prompt {cli::step} | {cli::pb_bar} {cli::pb_percent} [{cli::pb_eta}]"
  # )
  pb <- TRUE
  resps[which(!exists)] <- httr2::req_perform_parallel(reqs[!exists], 
                                                       paths = paths[!exists], 
                                                       on_error = on_error, 
                                                       progress = pb, 
                                                       max_active = max_active)
  resps[which(exists)] <- paths[exists]
  return(resps)
}
start <- Sys.time() # note start time for later
```

## Data

```{r}
prompts <- read_csv("input/prompts.csv", col_types = "cc")
nouns <- read_csv("input/nouns.csv", col_types = "icccccccc") |> 
  pivot_longer(cols = starts_with("Noun_"), names_to = "lang", values_to = "noun") |> 
  mutate(lang = str_remove(lang, "Noun_")) |> 
  rename(noun_id = id)
models <- c(
  "deepseek-r1:1.5b",
  "deepseek-r1:32b",
  "deepseek-r1:70b",
  "deepseek-r1:7b",
  "deepseek-r1:8b",
  "gemma:2b",
  "gemma:7b",
  "gemma3:12b",
  "gemma3:1b",
  "gemma3:4b",
  "gpt-oss:120b",
  "gpt-oss:20b",
  "llama3:8b",
  "llama3.1:70b",
  "llama3.1:8b",
  "llama3.2:1b",
  "llama3.2:3b",
  "mistral",
  "qwen2.5:0.5b",
  "qwen2.5:1.5b",
  "qwen2.5:3b",
  "qwen2.5:7b"
)
check_model_installed(models)
```

# Main

Construct the full set of prompts:

```{r}
all_prompts <- nouns |> 
  left_join(prompts, by = c("lang" = "version")) |> 
  mutate(prompt = paste(prompt, noun, sep = " "),
         schema = map(answers, function(a) {
           a <- str_split(a, ";")[[1]]
           list(type = "object", 
                properties = list(
                  gender = list(
                    type = "string", 
                    enum = a
                  )
                ),
                required = "gender"
           )
         })) |> 
  expand_grid(model = models) |> 
  mutate(prompt_id = paste(lang, noun, model, sep = "_"), .before = 1L) |> 
  # arrange by model to make sure no unnecessary loading and unloading of models happen
  arrange(model)
```

```{r}
dir.create("cache", showWarnings = FALSE)
results <- all_prompts |> 
  mutate(reqs = pmap(list(prompt, model, schema), function(p, m, s) {
    query(q = p,
          model = m,
          server = servers,
          format = s,
          output = "httr2_request")[[1]]
  })) |> 
  mutate(
    path = paste0("cache/", prompt_id, ".json"),
    resps = req_perform_parallel2(reqs, paths = path)
  )
```

```{r}
results |> 
  mutate(answer = map_chr(path, \(p) {
    jsonlite::read_json(p)[["message"]][["content"]] |> 
      jsonlite::fromJSON() |> 
      toString()
  })) |> 
  group_by(lang) |> 
  count(answer, sort = TRUE)
```



# wrap up

Save data if it takes long to preprocess it here. 
Afterwards we get some information which is important to reproduce the report.

```{r}
sessionInfo()
Sys.time()
# note how long the script takes to (re-)run
Sys.time() - start
```
