---
title: 1._prompting.qmd
author: Johannes B. Gruber
date: today
format:
  html:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

# Introduction

This file does all the prompting for the article "Gender Representation in Large Language Models: A Cross-Linguistic and Cross-Model Analysis"

## Packages

We use Ollama with the R package rollama for the prompting.
To make the prompting more efficient and make sure errors can be easily recovered from, I use a custom version of `httr2::req_perform_parallel()`, which rollama uses under the hood.

```{r}
#| include: false
library(tidyverse)
library(rollama)

# the script can use several ollama servers and distribute requests between them
# options(rollama_server = c("http://localhost:11434", "http://192.168.2.45:11434"))
if (!all(ping_ollama())) {
  stop("not all server are running")
}

# special req perform with cache and load balancing
req_perform_parallel2 <- function(reqs,
                                  paths = NULL,
                                  on_error = "continue",
                                  progress = TRUE,
                                  max_active = 10) {
  exists <- file.exists(paths)
  resps <- vector("list", length = length(reqs))
  if (progress) progress <- list(
    clear = TRUE,
    format = "Prompting {cli::pb_bar} | {cli::pb_percent} [{cli::pb_current}/{cli::pb_total} | rate: {cli::pb_rate} | ETA: {cli::pb_eta}]",
    format_done = "Done running {length(reqs)} prompts!"
  )

  resps[which(!exists)] <- httr2::req_perform_parallel(
    reqs[!exists], 
    paths = paths[!exists], 
    on_error = on_error, 
    progress = progress, 
    max_active = max_active
  ) |> 
    # I'm only extracting the body, which contains the path to the json file
    # to make this compatible with the other output
    map_chr("body")
  resps[which(exists)] <- paths[exists]
  return(resps)
}


parse_json_response <- function(r) {
    out <- ""
    if (!file.exists(r)) {
      cli::cli_warn("{r} does not exist")
      return(out)
    }
    res <- jsonlite::read_json(r)
    if (pluck_exists(res, "error")) {
      cli::cli_warn("Result contains error and will be deleted. Error:\n{res$error}")
      file.remove(r)
    } else {
      out <- pluck(res, "message", "content")
      if (str_detect(out, fixed("{"))) {
        out <- jsonlite::fromJSON(out)
      }
      out <- toString(out)
    }
    return(out)
  }
start <- Sys.time() # note start time for later
```

## Data

We construct a grid of the different prompts (one for each language), the selected nouns, and a range of different models.

```{r}
prompts <- read_csv("input/prompts.csv", col_types = "cc")
nouns <- read_csv("input/nouns.csv", col_types = "icccccccc") |> 
  pivot_longer(cols = starts_with("Noun_"), names_to = "lang", values_to = "noun") |> 
  mutate(lang = str_remove(lang, "Noun_")) |> 
  rename(noun_id = id)
models <- c(
  "deepseek-r1:1.5b",
  "deepseek-r1:32b",
  "deepseek-r1:70b",
  "deepseek-r1:7b",
  "deepseek-r1:8b",
  "gemma:2b",
  "gemma:7b",
  "gemma3:12b",
  "gemma3:1b",
  "gemma3:4b",
  "gpt-oss:120b",
  "gpt-oss:20b",
  "llama3:8b",
  "llama3.1:70b",
  "llama3.1:8b",
  "llama3.2:1b",
  "llama3.2:3b",
  "llama4:16x17b",
  "mistral",
  "qwen2.5:0.5b",
  "qwen2.5:1.5b",
  "qwen2.5:3b",
  "qwen2.5:7b"
)
# only use installed models
if (TRUE) {
  models <- intersect(rollama::list_models()$model, models)
}
check_model_installed(models)
```


# Main

We construct the full set of prompts here.
To get only valid answers, we use the structured output technique, to limit the choices the model can make.
We also run each combination of prompt and model 10 times, to add variation.
In previous iterations, we noticed that Ollama does some caching of replies.
So me make sure to first run all prompts in iteration 1, then all prompts in iteration 2 and so on.

```{r}
all_prompts <- nouns |> 
  left_join(prompts, by = c("lang" = "version")) |> 
  filter(lang == "English") |> 
  mutate(prompt = paste(prompt, noun, sep = "\n"),
         schema = map(answers, function(a) {
           a <- str_split(a, ";")[[1]]
           list(type = "object", 
                properties = list(
                  association = list(
                    type = "string", 
                    enum = a
                  )
                ),
                required = "association"
           )
         })) |> 
  expand_grid(model = models, seed = 1:10, temperature = seq(0, 1, 0.2)) |> 
  mutate(prompt_id = tolower(paste(lang, noun, model, temperature, seed, sep = "_")), .before = 1L) |> 
  # for temperature 0, the seed does not change the outcome, so we don't need it
  filter(!(temperature == 0 & seed > 1L)) |> 
  # arrange by model to make sure no unnecessary loading and unloading of models happen
  # but also arrange by seed to not run same prompt + model back to back
  arrange(seed, temperature, model)
```

I use rollama to prepare the requests to the servers, which are then made with the custom req_perform_parallel2 function defined above.

```{r}
dir.create("cache", showWarnings = FALSE)
grid <- all_prompts |> 
  mutate(req = pmap(list(prompt, model, schema, seed, temperature), function(p, m, s, sd, temp) {
    query(
      q = p,
      model = m,
      format = s,
      output = "httr2_request",
      model_params = list(seed = sd, temperature = temp)
    )[[1]]
  })) |> 
  mutate(path = paste0("cache/", prompt_id, ".json"))
glimpse(grid)
```

In total, the grid contains `r nrow(grid)` requests to Ollama

```{r}
results <- grid |> 
  mutate(resps = req_perform_parallel2(req, paths = path))
```

Results are written into a cache directory.
Once all prompting is done, we load the results from the cache directory and parse them.

```{r}
results_parsed <- results |> 
  mutate(answer = map_chr(resps, parse_json_response))
```

Quick glance at results:

```{r}
results_parsed |> 
  count(lang, answer, sort = TRUE)
```


# wrap up

save results:

```{r}
dir.create("data", showWarnings = FALSE)
saveRDS(results_parsed, "data/results_parsed.rds")
```

Some information about the session:

```{r}
sessionInfo()
Sys.time()
# note how long the script takes to (re-)run
Sys.time() - start
```
