---
title: 2. data cleaning
author: Johannes B. Gruber and Aleksandra Urman
date: today
format:
  html:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

# Introduction

This file does all the prompting for the article "Gender Representation in Large Language Models: A Cross-Linguistic and Cross-Model Analysis"

## Packages

We use Ollama with the R package rollama for the prompting.
To make the prompting more efficient and make sure errors can be easily recovered from, I use a custom version of `httr2::req_perform_parallel()`, which rollama uses under the hood.

```{r setup}
#| include: false
library(tidyverse)
apply_codebook <- function(string, codebook) {
  # remove all thinking first
  string <- str_remove(string, "<think>[\\S\\s]+</think>")
  hits <- map(codebook, function(pattern) {
    pattern <- paste0(pattern, collapse = "|")
    str_extract_all(string, regex(pattern, ignore_case = TRUE))[[1]] |> 
      tolower() |> 
      unique()
  })
  list(
    hit = unlist(hits, use.names = FALSE),
    category = names(compact(hits))
  )
}
start <- Sys.time() # note start time for later
```

## Data

We construct a grid of the different prompts (one for each language), the selected nouns, and a range of different models.

```{r}
results <- readRDS("data/results_parsed.rds") |> 
  select(prompt_id, noun_id, Noun, Valence, lang, noun, model, seed, temperature, answer)
# dictionary for gender extraction
gender_patterns <- list(
  English = list(
    men = c("\\bmen\\b", "\\bMan\\b", "\\bman\\b", "\\bmale[s]?\\b", "\\bMale[s]?\\b", "\\bmasculin", "\\bMasculin"),
    women = c("\\bwomen\\b", "\\bWoman\\b", "\\bwoman\\b", "\\bfemale[s]?\\b", "\\bFemale[s]?\\b", "\\bfeminin", "\\bFeminin"),
    nonbinary = c("\\bNon-binary\\b")
  ),
  German = list(
    men = c("\\bMann\\b", "\\bmann\\b", "\\bMänn", "\\bmänn", "\\bmännlich"),
    women = c("\\bFrau\\b", "\\bfrau\\b", "\\bweiblich"),
    nonbinary = c("\\bnicht-binär\\b", "\\bnichtbinär\\b", "\\bdivers\\b")
  ),
  Serbian = list(
    men = c("\\bmuškar", "\\bMuškar", "\\bmušk", "\\bMušk"),
    women = c("\\bžen", "\\bŽen", "\\bžensk", "\\bŽensk"),
    nonbinary = c("\\bnebinarn")
  ),
  Russian = list(
    men = c("\\bмужчин", "\\bМужчин", "\\bмужск", "\\bМужск"),
    women = c("\\bженщин", "\\bЖенщин", "\\bженск", "\\bЖенск"),
    nonbinary = c("\\bнебинарн", "\\bгендерквир")
  ),
  Czech = list(
    men = c("\\bmuž", "\\bMuž", "\\bmužsk", "\\bMužsk"),
    women = c("\\bžen", "\\bŽen", "\\bžensk", "\\bŽensk"),
    nonbinary = c("\\bnebinárn", "\\bnonbinárn")
  ),
  Albanian = list(
    men = c("\\bburr", "\\bBurr", "\\bmashkull"),
    women = c("\\bgra", "\\bGra", "\\bfemër"),
    nonbinary = c("\\bjobinar\\b", "\\bjo\\s+binar\\b")
  )
)
```


# Main
## Extracting just the gender answers

```{r}
results_parsed <- results |> 
  mutate(res = map2(
    .x = answer, 
    .y = lang,
    .f = function(.x, .y) apply_codebook(.x, codebook = gender_patterns[[.y]]),
    .progress = interactive()
  )) |> 
  unnest_wider(res)  
```

## Result validation

```{r}
results_parsed |> 
  unnest_longer(category) |> 
  count(category, lang, sort = TRUE) |> 
  DT::datatable()
```



# wrap up

save results:

```{r}
dir.create("data", showWarnings = FALSE)
saveRDS(results_parsed, "data/results_final.rds")
results_parsed |>
  rowwise() |> 
  mutate(
    hit = toString(hit),
    category = paste(category, collapse = "_")
  ) |> 
  rio::export("data/results_final.csv")
```

Some information about the session:

```{r}
sessionInfo()
Sys.time()
# note how long the script takes to (re-)run
Sys.time() - start
```
