---
title: 2. data cleaning
author: Johannes B. Gruber and Aleksandra Urman
date: today
format:
  html:
    toc: true
toc_depth: 3
toc_float: true
number_sections: true
---
  
# Introduction
  
This file cleans the raw data by applying a simply dictionary of the main categories we analyse in "Gender Representation in Large Language Models: A Cross-Linguistic and Cross-Model Analysis"

## Packages


```{r setup}
#| include: false
library(tidyverse)
apply_codebook <- function(string, codebook) {
  # remove all thinking first
  string <- str_remove(string, "<think>[\\S\\s]+</think>")
  hits <- map(codebook, function(pattern) {
    pattern <- paste0(pattern, collapse = "|")
    str_extract_all(string, regex(pattern, ignore_case = TRUE))[[1]] |> 
      tolower() |> 
      unique()
  })
  list(
    hit = unlist(hits),
    category = unique(names(unlist(hits)))
  )
}
start <- Sys.time() # note start time for later
```

## Data

We construct a grid of the different prompts (one for each language), the selected nouns, and a range of different models.

```{r}
results <- readRDS("data/results_parsed.rds") |> 
  select(prompt_id, noun_id, Noun, Valence, lang, noun, model, seed, temperature, answer)

```
```{r}
# dictionary for gender extraction
gender_patterns <- list(
  English = list(
    men = c(
      "\\bmen\\b", "\\bMan\\b", "\\bman\\b", "\\bMen\\b",
      "\\bmale[s]?\\b", "\\bMale[s]?\\b", 
      "\\bmasculin", "\\bMasculin",
      "\\bmanhood\\b", "\\bManhood\\b",
      "\\bmanliness\\b", "\\bManliness\\b"
    ),
    women = c(
      "\\bwomen\\b", "\\bWoman\\b", "\\bwoman\\b", "\\bWomen\\b",
      "\\bfemale[s]?\\b", "\\bFemale[s]?\\b", 
      "\\bfeminin", "\\bFeminin",
      "\\bwomanhood\\b", "\\bWomanhood\\b",
      "\\bwomanliness\\b", "\\bWomanliness\\b"
    ),
    nonbinary = c(
      "\\bNon-binary\\b", "\\bnon-binary\\b", "\\bnonbinary\\b", "\\bNonbinary\\b"
    )
  ),
  
  German = list(
    men = c(
      "\\bMann\\b", "\\bmann\\b", "\\bMänner\\b", "\\bmänner\\b",
      "\\bMänn", "\\bmänn",
      "\\bmännlich", "\\bMännlich",
      "\\bMännlichkeit\\b", "\\bmännlichkeit\\b",
      "\\bMannhaftigkeit\\b", "\\bmannhaftigkeit\\b"
    ),
    women = c(
      "\\bFrau\\b", "\\bfrau\\b", "\\bFrauen\\b", "\\bfrauen\\b",
      "\\bweiblich", "\\bWeiblich",
      "\\bWeiblichkeit\\b", "\\bweiblichkeit\\b",
      "\\bFraulichkeit\\b", "\\bfraulichkeit\\b"
    ),
    nonbinary = c(
      "\\bnicht-binär\\b", "\\bNicht-binär\\b",
      "\\bnichtbinär\\b", "\\bNichtbinär\\b",
      "\\bdivers\\b", "\\bDivers\\b",
      "\\bDiversen\\b", "\\bdiversen\\b",
      "\\bdiverse\\s+Personen\\b", "\\bDiverse\\s+Personen\\b",
      "\\bdiverse\\s+personen\\b",
      "\\bdiversen\\s+Personen\\b", "\\bDiversen\\s+Personen\\b",
      "\\bdiversen\\s+personen\\b"
    )
  ),
  
  Serbian = list(
    men = c(
      "\\bmuškar", "\\bMuškar",
      "\\bmuškarca\\b", "\\bMuškarca\\b",
      "\\bmuškarcima\\b", "\\bMuškarcima\\b",
      "\\bmušk", "\\bMušk",
      "\\bmuževnost", "\\bMuževnost",
      "\\bmuškost", "\\bMuškost"
    ),
    women = c(
      "\\bžen", "\\bŽen",
      "\\bžena\\b", "\\bŽena\\b",
      "\\bženama\\b", "\\bŽenama\\b",
      "\\bžensk", "\\bŽensk",
      "\\bženstvenost", "\\bŽenstvenost"
    ),
    nonbinary = c(
      "\\bnebinarn", "\\bNebinarn"
    )
  ),
  
  Russian = list(
    men = c(
      "\\bмужчин", "\\bМужчин",
      "\\bмужчина\\b", "\\bМужчина\\b",
      "\\bмужчины\\b", "\\bМужчины\\b",
      "\\bмужск", "\\bМужск",
      "\\bмужского\\b", "\\bМужского\\b",
      "\\bмужественность", "\\bМужественность",
      "\\bмужественност", "\\bМужественност",
      "\\bмужество\\b", "\\bМужество\\b"
    ),
    women = c(
      "\\bженщин", "\\bЖенщин",
      "\\bженщина\\b", "\\bЖенщина\\b",
      "\\bженщины\\b", "\\bЖенщины\\b",
      "\\bженск", "\\bЖенск",
      "\\bженского\\b", "\\bЖенского\\b",
      "\\bженственность", "\\bЖенственность",
      "\\bженственност", "\\bЖенственност"
    ),
    nonbinary = c(
      "\\bнебинарн", "\\bНебинарн",
      "\\bгендерквир", "\\bГендерквир"
    )
  ),
  
  Czech = list(
    men = c(
      "\\bmuž", "\\bMuž",
      "\\bmuži\\b", "\\bMuži\\b",
      "\\bmužů\\b", "\\bMužů\\b",
      "\\bmužsk", "\\bMužsk",
      "\\bmužského\\b", "\\bMužského\\b",
      "\\bmužnost", "\\bMužnost",
      "\\bmužství\\b", "\\bMužství\\b"
    ),
    women = c(
      "\\bžen", "\\bŽen",
      "\\bžena\\b", "\\bŽena\\b",
      "\\bženy\\b", "\\bŽeny\\b",
      "\\bžensk", "\\bŽensk",
      "\\bženského\\b", "\\bŽenského\\b",
      "\\bženství\\b", "\\bŽenství\\b",
      "\\bžennost", "\\bŽennost"
    ),
    nonbinary = c(
      "\\bnebinárn", "\\bNebinárn",
      "\\bnonbinárn", "\\bNonbinárn"
    )
  ),
  
  Albanian = list(
    men = c(
      "\\bburr", "\\bBurr",
      "\\bburra\\b", "\\bBurra\\b",
      "\\bburrave\\b", "\\bBurrave\\b",
      "\\bmashkull", "\\bMashkull",
      "\\bmashkuj\\b", "\\bMashkuj\\b",
      "\\bmashkullor", "\\bMashkullor",
      "\\bburrëri", "\\bBurrëri"
    ),
    women = c(
      "\\bgra\\b", "\\bGra\\b",
      "\\bgrave\\b", "\\bGrave\\b",
      "\\bgrua\\b", "\\bGrua\\b",
      "\\bgruaja\\b", "\\bGruaja\\b",
      "\\bfemër", "\\bFemër",
      "\\bfemra\\b", "\\bFemra\\b",
      "\\bfemëror", "\\bFemëror",
      "\\bgruar", "\\bGruar"
    ),
    nonbinary = c(
      "\\bjobinar", "\\bJobinar",
      "\\bjo\\s+binar", "\\bJo\\s+binar",
      "\\bjo-binar", "\\bJo-binar"
    )
  )
)
```

# Main
## Extracting just the gender answers

```{r}
results_parsed <- results |> 
  mutate(res = map2(
    .x = answer, 
    .y = lang,
    .f = function(.x, .y) apply_codebook(.x, codebook = gender_patterns[[.y]]),
    .progress = interactive()
  )) |> 
  unnest_wider(res)  
```
## Result empties

```{r}
empties<-results_parsed[results_parsed$hit=="NULL",]
print(summary(as.factor(empties$model)))
```
## Result validation

```{r}
results_parsed |> 
  unnest_longer(category) |> 
  count(category, lang, sort = TRUE) |> 
  DT::datatable()
```



# wrap up

save results:
  
```{r}
dir.create("data", showWarnings = FALSE)
saveRDS(results_parsed, "data/results_final.rds")
results_parsed |>
  rowwise() |> 
  mutate(
    hit = toString(hit),
    category = paste(category, collapse = "_")
  ) |> 
  rio::export("data/results_final.csv")
```

Some information about the session:
  
```{r}
sessionInfo()
Sys.time()
# note how long the script takes to (re-)run
Sys.time() - start
```
