---
title: 3. Analyses
author: Aleksandra Urman and Johannes B. Gruber 
date: today
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show the code"
    embed-resources: true
    toc_depth: 3
    toc_float: true
  docx: default
number_sections: true
---
  
# Introduction
  
This file contains the code to reproduce all figures, plots and in-text numbers in "Gender Representation in Large Language Models: A Cross-Linguistic and Cross-Model Analysis"

## Packages

```{r setup}
#| include: false
library(tidyverse)
library(infer)
library(DescTools)
library(knitr)
library(kableExtra)
opts_chunk$set(
  echo = knitr::is_html_output(),
  fig.width = 8.27,
  fig.height = 5.85,
  dpi = 300,
  comment = "#>"
)

# create folders for plots
dir.create("figures/png", recursive = TRUE, showWarnings = FALSE)
dir.create("figures/svg", recursive = TRUE, showWarnings = FALSE)

# Wrapper to suppress chi-square approximation warnings
# The "Chi-squared approximation may be incorrect" warning appears when expected
# cell counts are < 5. With our large sample sizes, the chi-square approximation
# remains valid despite low expected counts in some cells, so we suppress this
# specific warning.
chisq.test2 <- function(x, ...) {
  withCallingHandlers(
    chisq.test(x, ...),
    warning = function(w) {
      if (grepl("Chi-squared.approximation.may.be.incorrect", w)) {
        invokeRestart("muffleWarning")
      }
    }
  )
}

# Custom theme
paper_theme <- function(...) {
  font <- "Noto" # default CCR font
  theme_minimal(base_family = font) %+replace%
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(
        linetype = "dashed",
        linewidth = 0.15,
        colour = "#999999"
      ),
      plot.margin = margin(20, 20, 20, 20),

      # facets
      strip.background = element_rect(fill = "gray92", color = "grey70"),
      strip.text = element_text(
        family = font,
        size = 14,
        face = "bold",
        margin = margin(b = 5, t = 5)
      ),

      # fonts
      text = element_text(family = font, size = 10),
      legend.text = element_text(family = font, size = 8.5, colour = "grey30"),
      legend.title = element_text(
        family = font,
        size = 10,
        face = "bold",
        colour = "grey30"
      ),
      axis.text = element_text(family = font, size = 10),

      # legend
      legend.position = "bottom",
      legend.justification = "left",
      legend.direction = "horizontal"
    ) %+replace%
    theme(...)
}

theme_set(paper_theme())
update_geom_defaults("text", list(family = "Noto", size = 2.5))

# custom colours
gender_colours_main <- c(
  "Men only" = "#56B4E9",
  "Women only" = "#CC79A7",
  "Non-binary only" = "#E69F00",
  "Men + Women" = "#F0E442",
  "Women + Nonbinary" = "#D55E00",
  "Men + Nonbinary" = "#999999",
  "All three" = "#009E73"
)
scale_fill_paper <- function(...) {
  list(
    scale_fill_manual(
      values = gender_colours_main,
      guide = guide_legend(reverse = TRUE),
      ...
    )
  )
}

start <- Sys.time() # note start time for later
```

## Data

We construct a grid of the different prompts (one for each language), the selected nouns, and a range of different models.

```{r data}
results <- readRDS("data/results_final.rds")
categories <- read.csv("input/filtered_categories.csv") |>
  select(Noun = Noun_English, noun_category = Category)
```

```{r wrangling}
results_classified <- results |>
  mutate(
    Noun = tolower(Noun),
    gender_pattern = map_chr(category, function(cats) {
      if (is.null(cats)) {
        return("None/Unclear")
      }
      cats <- tools::toTitleCase(unique(cats))
      if (length(cats) == 1) {
        return(paste(cats, "only"))
      }
      if (length(cats) == 3) {
        return("All three")
      }
      return(paste(cats, collapse = " + "))
    }),
    gender_pattern = str_replace(
      gender_pattern,
      fixed("Nonbinary only"),
      "Non-binary only"
    ),
    gender_pattern_fct = fct_relevel(
      gender_pattern,
      rev(c(
        "Men only",
        "Women only",
        "Non-binary only",
        "Men + Women",
        "Men + Nonbinary",
        "Women + Nonbinary",
        "All three"
      ))
    )
  ) |>
  left_join(categories, by = "Noun")
```


# Main
## Table 1

| Model | English | German | Serbian | Russian | Czech | Albanian |
|-------|---------|--------|---------|---------|-------|----------|
| deepseek-r1:*[^deepseek-r1] | ✓ | - | - | - | - | - |
| gemma:2b/7b[^gemma2] | ✓ | - | - | - | - | - |
| gemma3:*[^gemma3] | ✓ | ✓ | - | - | - | - |
| gpt-oss:*[^gpt-oss] | ✓ | - | - | - | - | - |
| llama3:8b[^llama3] | ✓ | - | - | - | - | - |
| llama3.1:*[^llama31] | ✓ | ✓ | - | - | - | - |
| llama3.2:*[^llama32] | ✓ | ✓ | - | - | - | - |
| llama4:latest[^llama4] | ✓ | ✓ | - | - | - | - |
| mistral:latest[^mistral] | ✓ | ✓ | - | - | - | - |
| mistral-small3.1:latest[^mistral-small] | ✓ | ✓ | ✓ | ✓ | - | - |
| qwen2.5:*[^qwen] | ✓ | ✓ | - | ✓ | - | - |

: Official language support for LLMs used in this study. Based on systematic verification of official documentation from model providers (January 2025). *Legend*: ✓ = Officially supported with explicit documentation; - = Not officially supported or no explicit documentation available.

[^deepseek-r1]: Only English and Chinese officially supported ([GitHub](https://github.com/deepseek-ai/DeepSeek-R1); [Hugging Face](https://huggingface.co/deepseek-ai/DeepSeek-R1)).
[^gemma2]: Only English officially supported ([Model Card](https://ai.google.dev/gemma/docs/core/model_card)).
[^gemma3]: 140+ languages claimed, specific languages not listed ([Model Card](https://ai.google.dev/gemma/docs/core/model_card_3)).
[^gpt-oss]: Model card shows benchmarks for Arabic, Bengali, Chinese, French, German, Hindi, Indonesian, Italian, Japanese, Korean, Portuguese, Spanish, Swahili, and Yoruba. Serbian, Russian, Czech, and Albanian not explicitly listed ([Model Card](https://arxiv.org/pdf/2508.10925); also see [OpenAI announcement](https://openai.com/index/introducing-gpt-oss/); [GitHub](https://github.com/openai/gpt-oss)).
[^llama3]: "High-quality non-English data that covers over 30 languages", specific languages not listed ([Meta Blog](https://ai.meta.com/blog/meta-llama-3/)).
[^llama31]: Officially supports 8 languages (English, German, French, Italian, Portuguese, Hindi, Spanish, Thai), trained on a broader collection of languages than these 8 supported languages, specific languages not listed ([Model Card](https://github.com/meta-llama/llama-models/blob/main/models/llama3_1/MODEL_CARD.md)).
[^llama32]: Officially supports 8 languages (English, German, French, Italian, Portuguese, Hindi, Spanish, Thai), trained on a broader collection of languages than these 8 supported languages, specific languages not listed ([Model Card](https://github.com/meta-llama/llama-models/blob/main/models/llama3_2/MODEL_CARD.md)).
[^llama4]: Officially supports 12 languages (Arabic, English, French, German, Hindi, Indonesian, Italian, Portuguese, Spanish, Tagalog, Thai, Vietnamese), pre-training includes 200 total languages, specific languages not listed ([Model Card](https://github.com/meta-llama/llama-models/blob/main/models/llama4/MODEL_CARD.md)).
[^mistral]: Neither the official announcement nor the model card mention supported languages other than English ([model announcement](https://mistral.ai/news/announcing-mistral-7b); [Model Card](https://huggingface.co/mistralai/Mistral-7B-v0.1)).
[^mistral-small]: Explicitly languages *including* English, French, German, Greek, Hindi, Indonesian, Italian, Japanese, Korean, Malay, Nepali, Polish, Portuguese, Romanian, Russian, Serbian, Spanish, Swedish, Turkish, Ukrainian, Vietnamese, Arabic, Bengali, Chinese, Farsi ([Model Card](https://huggingface.co/mistralai/Mistral-Small-3.1-24B-Instruct-2503); [Mistral AI](https://mistral.ai/news/mistral-small-3-1)).
[^qwen]: 29+ languages officially supported, including Chinese, English, French, Spanish, Portuguese, German, Italian, Russian, Japanese, Korean, Vietnamese, Thai, and Arabic. Serbian, Czech, and Albanian not explicitly listed ([GitHub](https://github.com/QwenLM/Qwen2.5); [Model Card](https://www.alibabacloud.com/help/en/model-studio/what-is-qwen-llm)).

## Table 2

RQ1: Does gender representation differ across stereotype dimensions?

```{r tbl2}
#| tbl-cap: "Standardized Residuals from Chi-Square Test of Independence by Stereotype Dimension and Association Pattern"
df_rq1 <- results_classified |>
  filter(gender_pattern != "None/Unclear")
contingency_table <- table(df_rq1$noun_category, df_rq1$gender_pattern)
chi_test <- chisq.test2(contingency_table)
cramers_v <- CramerV(contingency_table)


# Statistics reported in-text
chi_square <- chi_test$statistic
df_chi <- chi_test$parameter
p_value <- chi_test$p.value
cramers_v_squared <- cramers_v^2
cat("Chi-square test of independence:\n")
cat(sprintf("χ²(%d) = %.3f, p %s\n", 
            df_chi, 
            chi_square, 
            ifelse(p_value < .001, "< .001", sprintf("= %.3f", p_value))))
cat(sprintf("Cramér's V = %.3f\n", cramers_v))
cat(sprintf("Cramér's V² = %.3f (effect size: %s)\n", 
            cramers_v_squared,
            ifelse(cramers_v_squared < .09, "small",
                   ifelse(cramers_v_squared < .25, "medium", "large"))))
#Reported as table
residuals_df <- as.data.frame.matrix(chi_test$stdres) |>
  rownames_to_column("Category") |>
  pivot_longer(
    -Category,
    names_to = "Gender_Pattern",
    values_to = "Std_Residual"
  ) |>
  mutate(
    Significance = case_when(
      abs(Std_Residual) > 3 ~ "***",
      abs(Std_Residual) > 2.58 ~ "**",
      abs(Std_Residual) > 1.96 ~ "*",
      TRUE ~ ""
    ),
    Display = sprintf("%.2f%s", Std_Residual, Significance)
  ) |>
  select("Stereotype Dimension" = Category, Gender_Pattern, Display) |>
  pivot_wider(names_from = Gender_Pattern, values_from = Display)
kable(residuals_df)
```


## Figure 1

```{r fig-1}
#| fig-cap: "Gender Association Patterns by Stereotype Dimension"
category_summary <- df_rq1 |>
  group_by(noun_category, gender_pattern = gender_pattern_fct) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(noun_category) |>
  mutate(
    percentage = (n / sum(n))
  ) |>
  # Keep main patterns for clarity
  filter(
    gender_pattern %in%
      c("Men only", "Women only", "Non-binary only", "All three", "Men + Women")
  )

category_summary_labeled <- category_summary |>
  group_by(noun_category) |>
  mutate(
    label_y = cumsum(percentage) - percentage / 2,
    label_text = ifelse(percentage > 3, paste0(round(percentage, 1), "%"), "")
  )

p1 <- ggplot(
  category_summary,
  aes(x = noun_category, y = percentage, fill = gender_pattern)
) +
  geom_col(
    position = position_dodge(width = 0.85),
    width = 0.8
  ) +
  geom_text(
    aes(
      y = percentage / 2,
      label = ifelse(
        percentage > 0.05,
        scales::percent(percentage, accuracy = 1),
        ""
      )
    ),
    position = position_dodge(width = 0.85),
    angle = 90,
    fontface = "bold",
    size = 5  
  ) +
  scale_y_continuous(
    labels = scales::percent,
    expand = expansion(mult = c(0, 0.12)),
    breaks = seq(0, 0.8, 0.2)
  ) +
  labs(
    x = "Stereotype Dimension",
    y = NULL,
    fill = "Gender Pattern"
  ) +
  guides(fill = guide_legend(nrow = 1)) +
  paper_theme(
    panel.grid.major.x = element_blank(),
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 14),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 13)
  ) +
  scale_fill_paper()

ggsave(
  "figures/png/figure_1.png",
  p1,
  width = 210,
  height = 148.5,
  units = "mm",
  dpi = 300
)
ggsave(
  "figures/svg/figure_1.svg",
  p1,
  width = 210,
  height = 148.5,
  units = "mm",
  dpi = 300
)
print(p1)
```

## Table 3

RQ2: Does gender representation vary across languages?

```{r rq2-filter}
# we remove results that do not correspond to a predefined noun category
df_rq2 <- results_classified |>
  filter(gender_pattern != "None/Unclear", !is.na(noun_category), !is.na(lang))
```

Sample size: `r nrow(df_rq2)`

::: {.callout-tip title="Detailed results" collapse="true"}
```{r tbl-3-calc}
# Three-way chi-square
contingency_3way <- table(
  df_rq2$lang,
  df_rq2$noun_category,
  df_rq2$gender_pattern
)
flat_table <- ftable(contingency_3way)
chi_3way <- chisq.test2(flat_table)
cramers_v_3way <- CramerV(flat_table)

#Reported in text
chi_square_3way <- chi_3way$statistic
df_3way <- chi_3way$parameter
p_value_3way <- chi_3way$p.value
cramers_v_squared_3way <- cramers_v_3way^2

# Print overall three-way results
cat("Three-way chi-square test:\n")
cat(sprintf("χ²(%d) = %.3f, p %s\n", 
            df_3way, 
            chi_square_3way, 
            ifelse(p_value_3way < .001, "< .001", sprintf("= %.3f", p_value_3way))))
cat(sprintf("Cramér's V = %.3f\n", cramers_v_3way))
cat(sprintf("Cramér's V² = %.3f (effect size: %s)\n\n", 
            cramers_v_squared_3way,
            ifelse(cramers_v_squared_3way < .09, "small",
                   ifelse(cramers_v_squared_3way < .25, "medium", "large"))))


category_results <- df_rq2 |>
  nest_by(noun_category) |>
  mutate(
    contingency = list(table(data$lang, data$gender_pattern)),
    chi_test = list(chisq.test2(contingency)),
    chi_square = chi_test$statistic,
    df = chi_test$parameter,
    p_value = chi_test$p.value,
    cramers_v = CramerV(contingency)
  ) |>
  ungroup()

# Print detailed results
category_results |>
  mutate(
    output = sprintf(
      "%s: χ²(%d) = %.2f, p < .001, Cramér's V = %.3f",
      noun_category,
      df,
      chi_square,
      cramers_v
    )
  ) |>
  pull(output) |>
  walk(~ cat(.x, "\n"))
cat("\n")
```

:::

```{r tbl-3}
#| tbl-cap: "Language Differences in Gender Association Patterns by Stereotype Dimension"
category_results |>
  select(
    Dimension = noun_category,
    Chi_square = chi_square,
    df,
    p_value
  ) |>
  mutate(
    Chi_square = sprintf("%.2f", Chi_square),
    p_value = "< .001",
    Cramers_V = sprintf("%.3f", category_results$cramers_v)
  ) |>
  arrange(desc(as.numeric(Cramers_V))) |>
  kable()
```

## Figure 2

```{r}
#| fig-cap: "Gender Association Patterns by Language and Stereotype Dimension"
lang_cat_grouped <- df_rq2 |>
  group_by(lang, noun_category, gender_pattern) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(lang, noun_category) |>
  mutate(
    total = sum(n),
    percentage = n / total
  ) |>
  filter(gender_pattern %in% c("Men only", "Women only", "Non-binary only"))

p2 <- ggplot(
  lang_cat_grouped,
  aes(x = percentage, y = fct_rev(noun_category), fill = gender_pattern)
) +
  geom_col(position = position_dodge(width = 0.85), width = 0.8) +
  geom_text(
    aes(
      x = percentage / 2,
      label = ifelse(
        percentage > 0.05,
        scales::percent(percentage, accuracy = 1),
        ""
      )
    ),
    colour = "grey10",
    position = position_dodge(width = 0.85),
    fontface = "bold"
  ) +
  facet_wrap(~lang, nrow = 2) +
  scale_x_continuous(
    labels = scales::percent,
    expand = expansion(mult = c(0, 0.12)),
    breaks = seq(0, 0.80, 0.20)
  ) +
  labs(
    x = NULL,
    y = NULL,
    fill = "Gender Pattern"
  ) +
  paper_theme(
    panel.grid.major.y = element_blank(),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13)
  ) +
  scale_fill_paper()

ggsave(
  "figures/png/figure_2.png",
  p2,
  width = 210,
  height = 148.5,
  units = "mm",
  dpi = 300
)
ggsave(
  "figures/svg/figure_2.svg",
  p2,
  width = 210,
  height = 148.5,
  units = "mm",
  dpi = 300
)
print(p2)
```

## Figure 3

RQ3a: Do different models vary in gender representation?

```{r rq3-filter}
df_rq3a <- results_classified |>
  filter(
    gender_pattern != "None/Unclear",
    !is.na(noun_category),
    !is.na(lang)
  ) |>
  mutate(
    model_family = case_when(
      str_detect(model, "gemma") ~ "Gemma",
      str_detect(model, "qwen") ~ "Qwen",
      str_detect(model, "llama") ~ "Llama",
      str_detect(model, "mistral") ~ "Mistral",
      str_detect(model, "phi") ~ "Phi",
      str_detect(model, "deepseek") ~ "DeepSeek",
      TRUE ~ "Other"
    )
  )

# For results reported in-text
contingency_model <- table(df_rq3a$model, df_rq3a$gender_pattern)
chi_model <- chisq.test(contingency_model)
cramers_v_model <- CramerV(contingency_model)
chi_square_model <- chi_model$statistic
df_model <- chi_model$parameter
p_value_model <- chi_model$p.value
cramers_v_squared_model <- cramers_v_model^2

cat("Chi-square test: Model × Gender Pattern\n")
cat(sprintf("χ²(%d) = %.3f, p %s\n", 
            df_model, 
            chi_square_model, 
            ifelse(p_value_model < .001, "< .001", sprintf("= %.3f", p_value_model))))
cat(sprintf("Cramér's V = %.3f\n", cramers_v_model))
cat(sprintf("Cramér's V² = %.3f (effect size: %s)\n\n", 
            cramers_v_squared_model,
            ifelse(cramers_v_squared_model < .09, "small",
                   ifelse(cramers_v_squared_model < .25, "medium", "large"))))

```

Sample size: `r nrow(df_rq3a)`

```{r}
#| tbl-cap: "Extra: Within-family variation"
# Within-family variation
family_consistency <- df_rq3a |>
  group_by(model_family, model, gender_pattern) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(model_family, model) |>
  mutate(total = sum(n), percentage = (n / total) * 100) |>
  group_by(model_family, gender_pattern) |>
  summarise(
    n_models = n(),
    mean_pct = mean(percentage),
    sd_pct = sd(percentage),
    min_pct = min(percentage),
    max_pct = max(percentage),
    range = max_pct - min_pct,
    .groups = "drop"
  ) |>
  arrange(model_family, desc(mean_pct))

family_summary <- family_consistency |>
  group_by(model_family) |>
  summarise(
    avg_sd = mean(sd_pct, na.rm = TRUE),
    max_range = max(range, na.rm = TRUE)
  ) |>
  arrange(desc(avg_sd))

kable(family_summary)
```

```{r}
#| fig-cap: "Distributions of gender association patterns by model"
model_plot_data <- df_rq3a |>
  mutate(
    gender_pattern = fct_relevel(
      gender_pattern,
      rev(c(
        "Men only",
        "Women only",
        "Non-binary only",
        "Men + Women",
        "Men + Nonbinary",
        "Women + Nonbinary",
        "All three"
      ))
    )
  ) |>
  group_by(model, gender_pattern) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(model) |>
  mutate(
    total = sum(n),
    percentage = (n / total)
  )

model_plot_labeled <- model_plot_data |>
  arrange(model, desc(gender_pattern)) |>
  group_by(model) |>
  mutate(
    label_x = cumsum(percentage) - percentage / 2,
    label_text = ifelse(
      percentage > 0.05,
      scales::percent(percentage, accuracy = 1),
      ""
    )
  )

p3 <- ggplot(
  model_plot_labeled,
  aes(x = percentage, y = model, fill = gender_pattern, label = label_text)
) +
  geom_col(position = "stack") +
  scale_x_continuous(labels = scales::percent) +
  geom_text(
    aes(x = label_x),
    size = 2.5,
    colour = "black",
    fontface = "bold"
  ) +
  labs(
    x = NULL,
    y = NULL,
    fill = "Gender Pattern"
  ) +
  scale_fill_paper() +
  guides(fill = guide_legend(nrow = 2, byrow = TRUE)) +
  paper_theme(
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13)
  )

ggsave(
  "figures/png/figure_3.png",
  p3,
  width = 300,
  height = 148.5,
  units = "mm",
  dpi = 300
)
ggsave(
  "figures/svg/figure_3.svg",
  p3,
  width = 300,
  height = 148.5,
  units = "mm",
  dpi = 300
)
print(p3)


#Faceted version of the plot
p3_faceted <- ggplot(
  model_plot_labeled,
  aes(x = percentage, y = model, fill = gender_pattern, label = label_text)
) +
  geom_col(position = "stack") +
  scale_x_continuous(labels = scales::percent) +
  geom_text(
    aes(x = 0.01),
    size = 2.5,
    colour = "black",
    fontface = "bold",
    hjust = 0
  ) +
  facet_wrap(~gender_pattern, nrow = 2) +
  labs(
    x = NULL,
    y = NULL,
    fill = "Gender Pattern"
  ) +
  scale_fill_paper() +
  guides(fill = guide_legend(nrow = 2, byrow = TRUE)) +
  paper_theme(
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13),
    plot.margin = margin(5, 20, 5, 5, "mm")
  )

ggsave(
  "figures/png/figure_3_faceted.png",
  p3_faceted,
  width = 400,
  height = 250,
  units = "mm",
  dpi = 300
)
ggsave(
  "figures/svg/figure_3_faceted.svg",
  p3_faceted,
  width = 400,
  height = 250,
  units = "mm",
  dpi = 300
)
print(p3_faceted)
```



## Supplementary Analysis: Variability Across Gender Patterns

This section is not reported in the main text but provides additional insights into pattern consistency across models.

```{r}
#| tbl-cap: "Supplementary: Coefficient of Variation by Gender Pattern"
# Calculate coefficient of variation for each pattern
pattern_variability <- model_plot_data |>
  group_by(gender_pattern) |>
  summarise(
    n_models = n(),
    mean_pct = mean(percentage),
    sd_pct = sd(percentage),
    min_pct = min(percentage),
    max_pct = max(percentage),
    range = max_pct - min_pct,
    cv = (sd_pct / mean_pct) * 100, # Coefficient of variation
    .groups = "drop"
  ) |>
  arrange(desc(cv))

cat("Variability metrics by gender pattern:\n")
cat(
  "(CV = Coefficient of Variation; higher values indicate greater variability)\n\n"
)
kable(pattern_variability, digits = 2)
```

```{r}
#| tbl-cap: "Supplementary: Variance by Gender Pattern"
# Test for homogeneity of variance
variance_comparison <- model_plot_data |>
  group_by(gender_pattern) |>
  summarise(
    variance = var(percentage),
    .groups = "drop"
  ) |>
  arrange(desc(variance))

kable(variance_comparison, digits = 2)
```

```{r}
#| tbl-cap: "Supplementary: Homogeneity Comparison for Men-only vs Women-only Patterns"
# Specific comparison: Women-only vs Men-only homogeneity
homogeneity_compare <- model_plot_data |>
  filter(gender_pattern %in% c("Women only", "Men only")) |>
  group_by(gender_pattern) |>
  summarise(
    mean = mean(percentage),
    sd = sd(percentage),
    cv = (sd(percentage) / mean(percentage)) * 100,
    range = max(percentage) - min(percentage),
    .groups = "drop"
  )

kable(homogeneity_compare, digits = 2)
```

**Interpretation:** Non-binary only shows highest variability (CV = `r sprintf("%.1f", pattern_variability$cv[pattern_variability$gender_pattern == "Non-binary only"])`%), compared to Women only (CV = `r sprintf("%.1f", pattern_variability$cv[pattern_variability$gender_pattern == "Women only"])`%) and Men only (CV = `r sprintf("%.1f", pattern_variability$cv[pattern_variability$gender_pattern == "Men only"])`%). This suggests Women-only and Men-only patterns are comparatively homogeneous across models.

## Table 4

RQ3b: Do models vary across languages?

```{r rq3b-filter}
df_rq3b <- results_classified |>
  filter(
    gender_pattern != "None/Unclear",
    !is.na(noun_category),
    !is.na(lang)
  )
```

Sample size: `r nrow(df_rq3b)`

::: {.callout-tip title="Detailed results" collapse="true"}
```{r tbl-4-calc}

cat("Within-language model consistency:\n")
cat(strrep("-", 60), "\n")

language_consistency <- df_rq3b |>
  group_by(lang, model, gender_pattern) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(lang, model) |>
  mutate(total = sum(n), percentage = (n / total) * 100) |>
  group_by(lang, gender_pattern) |>
  summarise(
    n_models = n(),
    mean_pct = mean(percentage),
    sd_pct = sd(percentage),
    min_pct = min(percentage),
    max_pct = max(percentage),
    range = max_pct - min_pct,
    .groups = "drop"
  ) |>
  arrange(lang, desc(mean_pct))

# Summary of consistency by language
language_summary <- language_consistency |>
  group_by(lang) |>
  summarise(
    avg_sd = mean(sd_pct, na.rm = TRUE),
    max_range = max(range, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(avg_sd)

# Print language consistency summary
cat("\nAverage standard deviation by language (lower = more consistent):\n")
language_summary |>
  mutate(
    output = sprintf("%s: Average SD = %.1f%%, Max Range = %.1f%%", 
                     lang, avg_sd, max_range)
  ) |>
  pull(output) |>
  walk(~ cat(.x, "\n"))

cat("\n")

# Display consistency table
kable(language_summary |>
        select("Language" = lang,
               "Average SD (%)" = avg_sd,
               "Max Range (%)" = max_range) |>
        mutate(across(where(is.numeric), ~round(.x, 1))),
      caption = "Within-Language Model Consistency")

# Three-way chi-square
contingency_3way <- table(df_rq3b$model, df_rq3b$lang, df_rq3b$gender_pattern)
flat_table <- ftable(contingency_3way)
chi_3way <- chisq.test2(flat_table)
cramers_v_3way <- CramerV(flat_table)

# Model effects within each language
language_results <- df_rq3b |>
  nest_by(lang) |>
  mutate(
    contingency = list(table(data$model, data$gender_pattern)),
    chi_test = list(chisq.test2(contingency)),
    chi_square = chi_test$statistic,
    df = chi_test$parameter,
    p_value = chi_test$p.value,
    cramers_v = CramerV(contingency)
  ) |>
  ungroup()

# Print detailed results
language_results |>
  mutate(
    output = sprintf(
      "%s: χ²(%d) = %.2f, p < .001, Cramér's V = %.3f",
      lang,
      df,
      chi_square,
      cramers_v
    )
  ) |>
  pull(output) |>
  walk(~ cat(.x, "\n"))
cat("\n")
```

:::

```{r tbl-4}
#| tbl-cap: "Model Effects within Each Language"
language_results |>
  select(
    Language = lang,
    Chi_square = chi_square,
    df,
    p_value
  ) |>
  mutate(
    Chi_square = sprintf("%.2f", Chi_square),
    p_value = "< .001",
    Cramers_V = sprintf("%.3f", language_results$cramers_v)
  ) |>
  arrange(desc(as.numeric(Cramers_V))) |>
  kable()
```

## Figure 4

```{r}
#| fig-cap: "The most frequent gender association patterns by model and language"
dominant_patterns <- df_rq3b |>
  group_by(lang, model, gender_pattern) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(lang, model) |>
  mutate(
    total = sum(n),
    percentage = (n / total)
  ) |>
  slice_max(percentage, n = 1) |>
  select(lang, model, dominant_pattern = gender_pattern, percentage)

p4 <- ggplot(
  dominant_patterns,
  aes(x = lang, y = model, fill = dominant_pattern)
) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(
    aes(label = scales::percent(percentage, accuracy = 1)),
    size = 2.5,
    colour = "black",
    fontface = "bold"
  ) +
  labs(
    x = NULL,
    y = NULL,
    fill = "Dominant Gender Pattern"
  ) +
  scale_fill_paper() +
  paper_theme(
    panel.grid.major = element_blank(),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13),
    axis.text.x = element_text(size = 13)
  ) +
  guides(fill = guide_legend(nrow = 2, byrow = TRUE))

ggsave(
  "figures/png/figure_4.png",
  p4,
  width = 250,
  height = 148.5,
  units = "mm",
  dpi = 300
)
ggsave(
  "figures/svg/figure_4.svg",
  p4,
  width = 250,
  height = 148.5,
  units = "mm",
  dpi = 300
)
print(p4)
```

## Table 5

RQ3c: Do models vary across stereotype dimensions?

```{r rq3c-filter}
df_rq3c <- results_classified |>
  filter(
    gender_pattern != "None/Unclear",
    !is.na(model),
    !is.na(noun_category)
  )
```

Sample size: `r nrow(df_rq3c)`

::: {.callout-tip title="Detailed results" collapse="true"}
```{r tbl-5-calc}
# Three-way chi-square
contingency_3way <- table(
  df_rq3c$model,
  df_rq3c$noun_category,
  df_rq3c$gender_pattern
)
flat_table <- ftable(contingency_3way)
chi_3way <- chisq.test2(flat_table)
cramers_v_3way <- CramerV(flat_table)
# Extract and report overall test statistics
chi_square_3way <- chi_3way$statistic
df_3way <- chi_3way$parameter
p_value_3way <- chi_3way$p.value
cramers_v_squared_3way <- cramers_v_3way^2

# Print overall three-way results
cat("Three-way chi-square test:\n")
cat(sprintf("χ²(%d) = %.3f, p %s\n", 
            df_3way, 
            chi_square_3way, 
            ifelse(p_value_3way < .001, "< .001", sprintf("= %.3f", p_value_3way))))
cat(sprintf("Cramér's V = %.3f\n", cramers_v_3way))
cat(sprintf("Cramér's V² = %.3f (effect size: %s)\n\n", 
            cramers_v_squared_3way,
            ifelse(cramers_v_squared_3way < .09, "small",
                   ifelse(cramers_v_squared_3way < .25, "medium", "large"))))
# Model effects within each category
category_results <- df_rq3c |>
  nest_by(noun_category) |>
  mutate(
    contingency = list(table(data$model, data$gender_pattern)),
    chi_test = list(chisq.test2(contingency)),
    chi_square = chi_test$statistic,
    df = chi_test$parameter,
    p_value = chi_test$p.value,
    cramers_v = CramerV(contingency)
  ) |>
  ungroup()

# Print detailed results
category_results |>
  mutate(
    output = sprintf(
      "%s: χ²(%d) = %.2f, p < .001, Cramér's V = %.3f",
      noun_category,
      df,
      chi_square,
      cramers_v
    )
  ) |>
  pull(output) |>
  walk(~ cat(.x, "\n"))
cat("\n")
```

:::

```{r tbl-5}
#| tbl-cap: "Model Effects within Each Stereotype Dimension"
category_results |>
  select(
    Dimension = noun_category,
    Chi_square = chi_square,
    df,
    p_value
  ) |>
  mutate(
    Chi_square = sprintf("%.2f", Chi_square),
    p_value = "< .001",
    Cramers_V = sprintf("%.3f", category_results$cramers_v)
  ) |>
  arrange(desc(as.numeric(Cramers_V))) |>
  kable()

cat("Within-dimension model consistency:\n")
cat(strrep("-", 60), "\n")

dimension_consistency <- df_rq3b |>  # or df_rq3a depending on your data structure
  group_by(noun_category, model, gender_pattern) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(noun_category, model) |>
  mutate(total = sum(n), percentage = (n / total) * 100) |>
  group_by(noun_category, gender_pattern) |>
  summarise(
    n_models = n(),
    mean_pct = mean(percentage),
    sd_pct = sd(percentage),
    min_pct = min(percentage),
    max_pct = max(percentage),
    range = max_pct - min_pct,
    .groups = "drop"
  ) |>
  arrange(noun_category, desc(mean_pct))

# Summary of consistency by stereotype dimension
dimension_summary <- dimension_consistency |>
  group_by(noun_category) |>
  summarise(
    avg_sd = mean(sd_pct, na.rm = TRUE),
    max_range = max(range, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(avg_sd)

# Print dimension consistency summary
cat("\nAverage standard deviation by stereotype dimension (lower = more consistent):\n")
dimension_summary |>
  mutate(
    output = sprintf("%s: Average SD = %.2f%%, Max Range = %.1f%%", 
                     noun_category, avg_sd, max_range)
  ) |>
  pull(output) |>
  walk(~ cat(.x, "\n"))

cat("\n")

# Display consistency table
kable(dimension_summary |>
        select("Stereotype Dimension" = noun_category,
               "Average SD (%)" = avg_sd,
               "Max Range (%)" = max_range) |>
        mutate(across(where(is.numeric), ~round(.x, 2))),
      caption = "Within-Dimension Model Consistency Across Gender Patterns")

```

## Figure 5

```{r}
#| fig-cap: "The most frequent gender association patterns by model and stereotype dimension."
dominant_patterns_cat <- df_rq3c |>
  group_by(noun_category, model, gender_pattern_fct) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(noun_category, model) |>
  mutate(
    total = sum(n),
    percentage = (n / total)
  ) |>
  slice_max(percentage, n = 1, with_ties = FALSE) |>
  select(
    noun_category,
    model,
    dominant_pattern = gender_pattern_fct,
    percentage
  )

p5 <- ggplot(
  dominant_patterns_cat,
  aes(x = noun_category, y = model, fill = dominant_pattern)
) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(
    aes(label = scales::percent(percentage, accuracy = 1)),
    size = 2.5,
    colour = "black",
    fontface = "bold"
  ) +
  labs(
    x = NULL,
    y = NULL,
    fill = "Dominant Gender Pattern"
  ) +
  paper_theme(
    panel.grid.major = element_blank(),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13),
    axis.text.x = element_text(size = 13)
  ) +
  scale_fill_paper()


ggsave(
  "figures/png/figure_5.png",
  p5,
  width = 250,
  height = 148.5,
  units = "mm",
  dpi = 300
)
ggsave(
  "figures/svg/figure_5.svg",
  p5,
  width = 250,
  height = 148.5,
  units = "mm",
  dpi = 300
)
print(p5)
```

## Table 6

RQ4: Do temperature effects vary across stereotype dimensions?

```{r rq4-filter}
df_rq4 <- results_classified |>
  filter(
    gender_pattern != "None/Unclear",
    !is.na(model),
    !is.na(lang),
    !is.na(temperature),
    !is.na(noun_category)
  )

temp_values <- sort(unique(df_rq4$temperature))
low_temp <- temp_values[1]
high_temp <- temp_values[length(temp_values)]
```

Sample size: `r nrow(df_rq4)`

::: {.callout-tip title="Detailed results" collapse="true"}
```{r tbl-6-calc}

# ============================================================================
# Pattern Stability Analysis: Most Frequent Pattern Across Temperatures
# ============================================================================

# Calculate most frequent pattern for each model-language-dimension-temperature combination
most_frequent_patterns <- df_rq4 |>
  group_by(noun_category, model, lang, temperature, gender_pattern) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(noun_category, model, lang, temperature) |>
  slice_max(n, n = 1, with_ties = FALSE) |>
  select(noun_category, model, lang, temperature, most_frequent = gender_pattern)

cat("Pattern Stability Analysis: Most Frequent Pattern Across Temperatures\n")
cat(strrep("=", 70), "\n\n")

# Check if the most frequent pattern is the same across all three temperatures
pattern_stability <- most_frequent_patterns |>
  group_by(noun_category, model, lang) |>
  summarise(
    n_temps = n_distinct(temperature),
    n_unique_patterns = n_distinct(most_frequent),
    stable = n_unique_patterns == 1,  # TRUE if same pattern across all temps
    patterns = paste(most_frequent[order(temperature)], collapse = ", "),
    .groups = "drop"
  )

# Calculate stability percentage by dimension
dimension_stability <- pattern_stability |>
  group_by(noun_category) |>
  summarise(
    total_combinations = n(),
    stable_combinations = sum(stable),
    stability_pct = (stable_combinations / total_combinations) * 100,
    .groups = "drop"
  ) |>
  arrange(desc(stability_pct))

cat("Stability by Stereotype Dimension:\n")
cat(strrep("-", 70), "\n")

dimension_stability |>
  mutate(
    output = sprintf("%s: %.1f%% stable (%d/%d model-language combinations)",
                     noun_category, stability_pct, stable_combinations, total_combinations)
  ) |>
  pull(output) |>
  walk(~ cat(.x, "\n"))

cat("\n")

# Display table
kable(dimension_stability |>
        select("Stereotype Dimension" = noun_category,
               "Total Combinations" = total_combinations,
               "Stable Combinations" = stable_combinations,
               "Stability %" = stability_pct) |>
        mutate(`Stability %` = round(`Stability %`, 1)),
      caption = "Most Frequent Pattern Stability Across Temperature Levels by Dimension")

# ============================================================================
# Stability by Language within Each Dimension
# ============================================================================

cat("\n\nStability by Language within Each Dimension:\n")
cat(strrep("=", 70), "\n\n")

language_dimension_stability <- pattern_stability |>
  group_by(noun_category, lang) |>
  summarise(
    total_combinations = n(),
    stable_combinations = sum(stable),
    stability_pct = (stable_combinations / total_combinations) * 100,
    .groups = "drop"
  ) |>
  arrange(noun_category, desc(stability_pct))

# Print by dimension
for (dim in unique(language_dimension_stability$noun_category)) {
  cat(sprintf("\n%s:\n", dim))
  cat(strrep("-", 70), "\n")
  
  language_dimension_stability |>
    filter(noun_category == dim) |>
    mutate(
      output = sprintf("  %s: %.1f%% stable (%d/%d)",
                       lang, stability_pct, stable_combinations, total_combinations)
    ) |>
    pull(output) |>
    walk(~ cat(.x, "\n"))
}

cat("\n")

# Full table
kable(language_dimension_stability |>
        select("Stereotype Dimension" = noun_category,
               "Language" = lang,
               "Total Combinations" = total_combinations,
               "Stable Combinations" = stable_combinations,
               "Stability %" = stability_pct) |>
        mutate(`Stability %` = round(`Stability %`, 1)),
      caption = "Most Frequent Pattern Stability by Dimension and Language")





# Calculate entropy by category and temperature
entropy_by_cat_temp <- df_rq4 |>
  group_by(noun_category, model, lang, temperature, gender_pattern) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(noun_category, model, lang, temperature) |>
  mutate(
    total = sum(n),
    proportion = n / total,
    entropy_component = -proportion * log(proportion)
  ) |>
  group_by(noun_category, model, lang, temperature) |>
  summarise(
    shannon_entropy = sum(entropy_component),
    normalized_entropy = shannon_entropy / log(7),
    .groups = "drop"
  )

entropy_by_cat_comparison <- entropy_by_cat_temp |>
  filter(temperature %in% c(low_temp, high_temp)) |>
  dplyr::select(noun_category, model, lang, temperature, normalized_entropy) |>
  pivot_wider(
    names_from = temperature,
    values_from = normalized_entropy,
    names_prefix = "temp_"
  ) |>
  mutate(
    low_temp_entropy = get(paste0("temp_", low_temp)),
    high_temp_entropy = get(paste0("temp_", high_temp)),
    entropy_change = high_temp_entropy - low_temp_entropy
  )

# ANCOVA: Temperature × Dimension
entropy_by_cat_temp$noun_category <- factor(entropy_by_cat_temp$noun_category)
ancova_model <- aov(
  normalized_entropy ~ temperature * noun_category,
  data = entropy_by_cat_temp
)

cat("ANCOVA: Temperature × Dimension\n")
print(summary(ancova_model))
cat("\n")
```

:::

```{r tbl-6}
#| tbl-cap: "Gender Parity Gap Statistics by Stereotype Dimension"
# Calculate parity gap for each model × dimension
model_dimension_parity <- df_rq3c |>
  group_by(model, noun_category, gender_pattern) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(model, noun_category) |>
  mutate(
    total = sum(n),
    percentage = (n / total) * 100
  ) |>
  filter(gender_pattern %in% c("Men only", "Women only")) |>
  pivot_wider(
    names_from = gender_pattern,
    values_from = percentage,
    values_fill = 0
  ) |>
  mutate(
    parity_gap = `Men only` - `Women only`
  )

# Summary by dimension
parity_variation <- model_dimension_parity |>
  group_by(noun_category) |>
  summarise(
    mean_parity_gap = mean(parity_gap),
    sd_parity_gap = sd(parity_gap),
    range = max(parity_gap) - min(parity_gap),
    .groups = "drop"
  ) |>
  arrange(desc(abs(mean_parity_gap)))

parity_variation |>
  mutate(
    across(where(is.numeric), ~ sprintf("%.1f", .x))
  ) |>
  kable(
    col.names = c(
      "Dimension",
      "Mean Gap (%)",
      "SD (%)",
      "Range (%)"
    )
  )
```

*Note. Parity gap = % Men-only − % Women-only associations. Positive values indicate male bias; negative values indicate female bias. SD = standard deviation across all 24 models. Range = maximum parity gap − minimum parity gap across models.*

## Figure 6

```{r}
#| fig-cap: "Gender parity gap by stereotype dimension"
p6 <- ggplot(
  parity_variation,
  aes(x = reorder(noun_category, mean_parity_gap), y = mean_parity_gap)
) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    colour = "grey50",
    linewidth = 1.2
  ) +
  geom_linerange(
    aes(
      ymin = mean_parity_gap - sd_parity_gap,
      ymax = mean_parity_gap + sd_parity_gap
    ),
    linewidth = 10,
    alpha = 0.3,
    colour = "steelblue"
  ) +
  geom_point(size = 7, colour = "steelblue") +
  geom_text(
    aes(label = sprintf("%+.1f%%", mean_parity_gap)),
    hjust = ifelse(parity_variation$mean_parity_gap > 0, -0.3, 1.3),
    size = 6.5,
    fontface = "bold",
    colour = "grey30"
  ) +
  annotate(
    "text",
    x = 5.5,
    y = 50,
    label = "Male-biased →",
    size = 7,
    family = "Noto",
    fontface = "italic",
    colour = "grey30"
  ) +
  annotate(
    "text",
    x = 5.5,
    y = -50,
    label = "← Female-biased",
    size = 7,
    family = "Noto",
    fontface = "italic",
    colour = "grey30"
  ) +
  coord_flip(ylim = c(-80, 80)) +
  labs(
    x = NULL,
    y = NULL,
    caption = "Large shaded regions indicate high model-to-model variation"
  ) +
  paper_theme(
    axis.text.x = element_text(size = 18),
    axis.text.y = element_text(size = 18),
    plot.caption = element_text(size = 13)
  )

ggsave(
  "figures/png/figure_6.png",
  p6,
  width = 210,
  height = 148.5,
  units = "mm",
  dpi = 300
)
ggsave(
  "figures/svg/figure_6.svg",
  p6,
  width = 210,
  height = 148.5,
  units = "mm",
  dpi = 300
)
print(p6)
```

*Note: Points show mean parity gap (% Men-only - % Women-only); shaded regions show ±1 SD across models. Large shaded regions indicate high model-to-model variation. Positive values indicate men male bias (above zero line); negative values indicate women female bias (below zero line).*

## Supplementary Analysis: Pattern Stability Across Temperatures

This section is not reported in the main text but examines whether models maintain consistent dominant patterns across different temperature settings.

```{r}
#| tbl-cap: "Supplementary: Pattern Stability by Stereotype Dimension"
# Calculate dominant pattern for each model/lang/temp/dimension combination
dominant_by_cat_temp <- df_rq4 |>
  group_by(noun_category, model, lang, temperature, gender_pattern) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(noun_category, model, lang, temperature) |>
  mutate(
    total = sum(n),
    percentage = (n / total) * 100
  ) |>
  slice_max(percentage, n = 1, with_ties = FALSE) |>
  dplyr::select(
    noun_category,
    model,
    lang,
    temperature,
    dominant_pattern = gender_pattern,
    dominance_strength = percentage
  )

# Check if patterns remain stable across temperatures
pattern_stability_by_cat <- dominant_by_cat_temp |>
  group_by(noun_category, model, lang) |>
  summarise(
    n_temps = n(),
    n_unique_patterns = n_distinct(dominant_pattern),
    is_stable = n_unique_patterns == 1,
    .groups = "drop"
  )

# Summary by dimension
stability_summary <- pattern_stability_by_cat |>
  group_by(noun_category) |>
  summarise(
    n_combos = n(),
    n_stable = sum(is_stable),
    pct_stable = round(100 * n_stable / n(), 1),
    .groups = "drop"
  ) |>
  arrange(desc(pct_stable))

kable(
  stability_summary,
  col.names = c("Dimension", "N Combinations", "N Stable", "% Stable")
)
```

```{r}
#| fig-cap: "Additional plot: Temperature by Mean Normalized Entropy"
entropy_cat_summary_viz <- entropy_by_cat_temp |>
  group_by(noun_category, temperature) |>
  summarise(
    mean_entropy = mean(normalized_entropy),
    sd_entropy = sd(normalized_entropy),
    .groups = "drop"
  )

p6_alt <- ggplot(
  entropy_cat_summary_viz,
  aes(x = temperature, y = mean_entropy)
) +
  geom_line(linewidth = 1.2, colour = "steelblue") +
  geom_point(size = 3, colour = "steelblue") +
  geom_ribbon(
    aes(ymin = mean_entropy - sd_entropy, ymax = mean_entropy + sd_entropy),
    alpha = 0.2,
    fill = "steelblue"
  ) +
  facet_wrap(~noun_category, nrow = 2) +
  scale_x_continuous(breaks = temp_values) +
  labs(
    x = "Temperature",
    y = "Mean Normalized Entropy"
  ) +
  paper_theme(panel.grid.major = element_blank()) +
  scale_fill_paper()
print(p6_alt)
```

## Table 7

```{r tbl-7}
#| tbl-cap: "Mean concentration strength by stereotype dimension for the lowest (0) and highest (1) temperature"
# Calculate dominant pattern strength (concentration)
dominant_by_cat_temp <- df_rq4 |>
  group_by(noun_category, model, lang, temperature, gender_pattern) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(noun_category, model, lang, temperature) |>
  mutate(
    total = sum(n),
    percentage = (n / total) * 100
  ) |>
  slice_max(percentage, n = 1, with_ties = FALSE) |>
  dplyr::select(
    noun_category,
    model,
    lang,
    temperature,
    dominant_pattern = gender_pattern,
    dominance_strength = percentage
  )

concentration_by_cat <- dominant_by_cat_temp |>
  filter(temperature %in% c(low_temp, high_temp)) |>
  dplyr::select(noun_category, model, lang, temperature, dominance_strength) |>
  pivot_wider(
    names_from = temperature,
    values_from = dominance_strength,
    names_prefix = "temp_"
  ) |>
  mutate(
    low_temp_value = get(paste0("temp_", low_temp)),
    high_temp_value = get(paste0("temp_", high_temp)),
    change = high_temp_value - low_temp_value
  )

concentration_summary <- concentration_by_cat |>
  group_by(noun_category) |>
  summarise(
    mean_low = mean(low_temp_value, na.rm = TRUE),
    mean_high = mean(high_temp_value, na.rm = TRUE),
    mean_change = mean(change, na.rm = TRUE),
    sd_change = sd(change, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(abs(mean_change)))

concentration_summary |>
  mutate(
    across(where(is.numeric), ~ sprintf("%.2f", .x))
  ) |>
  kable(
    col.names = c(
      "Dimension",
      "Mean Low (T=0)",
      "Mean High (T=1)",
      "Mean Change",
      "SD"
    )
  )
```

## Table 8

```{r tbl-8}
#| tbl-cap: "Shannon entropy values by stereotype dimension for the lowest (0) and highest (1) temperature"
# Simple entropy summary (calculated in Table 6 section)
entropy_by_cat_comparison |>
  group_by(noun_category) |>
  summarise(
    mean_low_entropy = mean(low_temp_entropy, na.rm = TRUE),
    sd_low_entropy = sd(low_temp_entropy, na.rm = TRUE),
    mean_high_entropy = mean(high_temp_entropy, na.rm = TRUE),
    sd_high_entropy = sd(high_temp_entropy, na.rm = TRUE),
    mean_entropy_change = mean(entropy_change, na.rm = TRUE),
    sd_entropy_change = sd(entropy_change, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(abs(mean_entropy_change)) |>
  select(
    noun_category,
    mean_low_entropy,
    mean_high_entropy,
    mean_entropy_change,
    sd_entropy_change
  ) |>
  mutate(
    across(where(is.numeric), ~ sprintf("%.3f", .x))
  ) |>
  kable(
    col.names = c(
      "Dimension",
      "Low Temp (T=0)",
      "High Temp (T=1)",
      "Change",
      "SD"
    )
  )
```

## Supplementary Analysis: Linear Regression by Dimension

This section is not reported in the main text but provides dimension-specific regression analyses of entropy changes with temperature.

```{r models}
#| tbl-cap: "Supplementary: Linear Regression Results by Stereotype Dimension"
#| output: "asis"
noun_categories <- entropy_by_cat_temp |>
  pull("noun_category") |>
  as.character() |>
  unique() |>
  sort() |>
  setNames(nm = _)
models <- map(noun_categories, function(cat_name) {
  cat_data <- entropy_by_cat_temp |> filter(noun_category == cat_name)
  lm(normalized_entropy ~ temperature, data = cat_data)
})
texreg::knitreg(models)
```


# wrap up

Some information about the session:
  
```{r}
sessionInfo()
Sys.time()
# note how long the script takes to (re-)run
Sys.time() - start
```
